---
title: "Louisville Police UOF Pre-processing and EDA"
output: html_document
---

The following document analyzes a use-of-force (UOF) report from the Louisville Metro Police to see how often force is used, how often it's violent in nature, how effective it is, and which officers are most at fault for justifiably (or unjustifiably) using violent force against minorities. 

__Data Source:__

* uof_louisville.csv

__Below is a list of variables that were of importance:__

* date_of_occurrence - the date the UOF incident occurred.
* time_of_occurrence - the time the UOF incident occurred.
* citizen_injured - whether the citizen was injured in the incident.
* service_rendered - how the police officer was called to the scene.
* reason_for_force - why force was used.
* force_used_{number} - the type of force used.
* force_used_{number}_effective - whether the force was effective.
* name_badge_number - the officer name and badge number.
* badge_number - only the badge number of the officer


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

#### Reading in UOF data

```{r}
uof <- read_csv('uof_louisville.csv')
```

#### Creating a new column which is the hour of the time_of_occurrence column, and finding the hour that is the most frequent for UOF reports

```{r}
frequent_hour <- uof %>%
  mutate(hour = lubridate::hour(time_of_occurrence)) %>%
  count(hour, sort = T) %>%
  slice(1) %>%
  pull(hour)
```

#### Creating a new column which is the month of the date_of_occurrence column, and finding the month with the least amount of UOF incidents

```{r}
least_frequent_month <- uof %>%
  mutate(month = lubridate::month(date_of_occurrence)) %>%
  count(month, sort = T) %>%
  slice(12) %>%
  pull(month)
```

#### Creating a new column named 'day' which is the day value of the date_of_occurrence column. 

#### Finding the day of the week that has the most number of UOF reports

```{r}
most_frequent_day <- uof %>%
  mutate(day = lubridate::wday(date_of_occurrence, label = TRUE)) %>%
  count(day, sort = T) %>%
  slice(1) %>%
  pull(day)
```

#### Creating a table to see if there is any relationship as to when UOF incidents happen most in the month

```{r}
day_distribution <- uof %>% count(day = lubridate::day(date_of_occurrence), sort = T)
day_distribution <- day_distribution %>% mutate(fraction = n / nrow(uof))
day_distribution <- janitor::adorn_totals(day_distribution)
```
```{r, echo=FALSE}
show(day_distribution)
```

#### Finding all unique categories of force_used_1 and force_used_2

```{r}
force_used_1 <- uof %>% distinct(force_used_1) %>% pull()
force_used_2 <- uof %>% distinct(force_used_2) %>% pull()
```

#### Extracting all the unique types of forces used from force_used_1...force_used_8

```{r}
all_force <- uof %>%
  distinct(force_used_1, force_used_2, force_used_3, force_used_4, force_used_5, force_used_6, force_used_7, force_used_8) %>%
  t() %>%
  c() %>%
  unique()
```

#### Defining a new vector which contains all the violent UOF tactics

```{r}
violent_force <- c("take down", "hobble", "ecw cartridge deployed", "knee strike(s)",
                   "12 ga. sock round", "take-down", "impact weapon",
                   "kick", "deadly force used")
```

#### Creating a column which is a binary variable equal to 1 if force_used_1 contains any values from violent_force and 0 otherwise

```{r}
uof <- uof %>%
  mutate(violent_uof_1 = ifelse(force_used_1 %in% violent_force ,1, 0))
```

#### Creating a table to see the counts and proportions of the services rendered that ended up in violent interactions 

```{r}
violent_force_service_table <- uof %>%
  filter(violent_uof_1 == 1) %>%
  count(service_rendered, sort = T) %>%
  mutate(fraction = n/nrow(filter(uof, violent_uof_1 == 1))) %>%
  janitor::adorn_totals()
```
```{r, echo=FALSE}
show(violent_force_service_table)
```

#### Filtering citizen_gender to only include males and females, excluding all NA values from citizen_race, and creating a new column which is binary 1 if force_used_1_effective is equal to 'yes' and 0 otherwise

```{r}
uof_filtered <- uof %>%
  filter(citizen_gender == 'male'| citizen_gender == 'female') %>%
  filter(!is.na(citizen_race)) %>%
  mutate(force_used_1_effective_binary = ifelse(force_used_1_effective == 'yes', 1, 0))
```

#### Creating table to group by gender and race to see how many times violent force was used on them, how often it was effective, and what fraction of the time it was effective

```{r}
uof_filtered_table <- uof_filtered %>%
  group_by(citizen_gender, citizen_race) %>%
  summarise(effective_1 = sum(force_used_1_effective_binary, na.rm = T), counts = n()) %>%
  janitor::adorn_totals() %>%
  mutate(fraction_effective = effective_1/counts)
```
```{r, echo=FALSE}
show(uof_filtered_table)
```

#### Finding which officers are involved the most in incidents, which officers have the most 'violent' interactions, and which officers have the highest percentage of violent interactions

```{r}
violent_officers <- uof %>%
  group_by(name_badge_number) %>%
  filter(force_used_1 != 'none' & !is.na(force_used_1)) %>%
  summarise(violent_amount = sum(violent_uof_1), counts = n()) %>%
  mutate(fraction_violent = violent_amount/counts) %>%
  arrange(desc(fraction_violent))
```
```{r, echo=FALSE}
show(violent_officers)
```

#### Finding which officers had the most violent interactions by the race of the citizens

```{r}
violent_officers_prej <- uof %>%
  group_by(name_badge_number, citizen_race) %>%
  summarise(violent_force_race = sum(violent_uof_1)) %>%
  arrange(desc(violent_force_race))
```
```{r, echo=FALSE}
show(violent_officers_prej)
```

#### Finding what reason cops were called that led to force

```{r}
force_by_reason <- uof %>%
  group_by(service_rendered) %>%
  filter(force_used_1 != 'none' & !is.na(force_used_1)) %>%
  count(sort = T)
```
```{r, echo=FALSE}
show(force_by_reason)
```

#### Finding what reason cops were called that led to violent interaction

```{r}
violent_by_reason <- uof %>%
  group_by(service_rendered) %>%
  summarise(violent = sum(violent_uof_1)) %>%
  arrange(desc(violent))

force_failed_violent <- uof %>%
  filter(force_used_1_effective == 'no' & violent_uof_1 == 1) %>%
  nrow()

violent_interaction <- uof %>%
  filter(violent_uof_1 == 1) %>%
  nrow()
```
```{r, echo=FALSE}
show(violent_by_reason)
```

#### Finding the failure rate of using violence in an interaction with a citizen

```{r}
violent_failure_rate <- force_failed_violent/violent_interaction
```
```{r, echo=FALSE}
show(violent_failure_rate)
```

#### Finding what the citizen influences were that led to use of both regular force and violent force

```{r}
citizen_influence_force <- uof %>%
  filter(force_used_1 != 'none' & !is.na(force_used_1)) %>%
  group_by(citizen_influence_assessment) %>%
  count() %>%
  arrange(desc(n))
citizen_influence_violent <- uof %>%
  filter(force_used_1 != 'none' & !is.na(force_used_1)) %>%
  group_by(citizen_influence_assessment) %>%
  summarise(violent_influnece = sum(violent_uof_1)) %>%
  arrange(desc(violent_influnece))
```
```{r, echo=FALSE}
show(citizen_influence_force)
show(citizen_influence_violent)
```

